# WEEK 6: Analysis of longitudinal data

Last but not least week! First I read in two data sets that I just wrangled to longitudinal form.

```{r, message=FALSE, warning=FALSE}
# Set working directory:
setwd("C:/Users/mikniemi/OneDrive - University of Helsinki/OpenDataScience/IODS/IODS-project")

# Access the necessary tool packages:
library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)

# Load, factor and look at the data sets:
BPRS <- read.table("Data/BPRS.txt")
BPRSL <- read.table("Data/BPRSL.txt")
BPRSL$treatment <- factor(BPRSL$treatment)
# Fix the data set such as every individual has its own subject number:
for(i in 1:nrow(BPRSL)){
  if(BPRSL[i,"treatment"] == 2){
    BPRSL[i,"subject"] <- BPRSL[i,"subject"] + 20
  }
}
BPRSL$subject <- factor(BPRSL$subject)
glimpse(BPRSL)

RATS <- read.table("Data/RATS.txt")
RATSL <- read.table("Data/RATSL.txt")
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)
glimpse(RATSL)
```

There are glimpses of these two data sets above. The first data set **BPRSL** contains weekly measures of 40 males assigned to two treatment groups (1 or 2). The data is used to evaluate patients suspected having schizophrenia. The lower value of BPRS (minimum 18) means the less symptoms of schizophrenia present.
The latter data set **RATSL** contains weekly weight measurements (in grams) of rats divided into three groups having different diets. In here we are interested if different diets effects to rats' growth.

### Summary Measure Approach using the RATSL data

```{r}
# Draw the plot:
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

**Figure 1.** Individual response profiles by diet group for the RATS data.

```{r}
# Standardise the variable Weight:
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()

# Plot again with the standardized weights:
ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$stdWeight), max(RATSL$stdWeight)))
```

**Figure 2.** The corresponding individual response profiles after standardization.

```{r, message=FALSE, warning=FALSE}
# Number of measurement times:
n <- RATSL$Time %>% unique() %>% length()

# Summary data with mean and standard error of Weight by Group and Time:
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data:
glimpse(RATSS)

# Plot the mean profiles:
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.5)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

**Figure 3.** Summarized mean response profiles of the three diet groups. Error bars illustrates the standard errors of mean.

```{r, message=FALSE, warning=FALSE}
# Create a summary data by Group and ID with mean as the summary variable (ignoring baseline WD1):
RATSLWDS <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data:
glimpse(RATSLWDS)

# Draw a boxplot of the mean versus treatment:
ggplot(RATSLWDS, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=3, fill = "white") +
  scale_y_continuous(name = "mean(Weight)")
```

**Figure 4.** Possible outliers of the three groups can be detected from the boxplots above.

```{r, message=FALSE, warning=FALSE}
# Add the baseline (WD1) from the original data as a new variable to the summary data:
RATSLWDS1 <- RATSLWDS %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response variable:
fit <- lm(mean ~ baseline + Group, data = RATSLWDS1)

# Compute the analysis of variance table for the fitted model with anova():
anova(fit)
```

**Table 1.** A variance table of Anova test, which shows us that the baseline is clearly significant component of the model, but the statistical significance of Group-variable is unclear. **The result means that we could gather the similar data by probability of 7.6 % unless diet has no effect on rat growth.**

### Linear Mixed Effects Models using using the BPRSL data

```{r}
# Plot the BPRSL data:
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(linetype = treatment, color = treatment)) +
  scale_x_continuous(name = "Weeks", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "BPRS", breaks = seq(0, 100, 20)) +
  theme(legend.position = "top")
```

**Figure 5.** Plot of BPRS indexes against time. Individuals having different treatments are illustrated by different line types and colors.

```{r}
# Create a regression model BPRS_reg and print out the model summary:
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRS_reg)
```

**Table 2.** The summary of simple linear regression model, where BPRS is modeled using time (week) and dummy variable of treatment. According to this print, treatment would not be significantly affecting to BPRS index value. **But remember that the use of linear regression assumes that all the measurements were independent, which is not true here due to the longitudinal character of the data.**

```{r}
# Create a random intercept model and print out the model summary:
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref)
```

**Table 3.** Now there is the summary of a *random intercept model*, which creates own intercept for every subject, but fits the same slope for every subject. There is large variance between the intercept as can be seen from the table (std. dev. of 6.9). However, in conclusion **the residual standard error is 10.2, which is significantly lower than the corresponding result from linear regression model: 12.4**

```{r}
# Create a random intercept and random slope model, and print out the summary:
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref1)
```

**Table 4.** The summary of *random intercept and random slope model* to the BPRS data. The residuals are a bit lower than corresponding results in the previous table.

```{r}
# Perform an ANOVA test on the two models:
anova(BPRS_ref1, BPRS_ref)
```

**Table 5.** ANOVA test of the two previous models. The latter model (ref1) fits the data a bit better (e.g. lower logLikelihood), but the difference is not clear. The AIC, which measures model parsimony, is almost equal. In conclusion, **random slope improves the model fit in the study data, but if we would like to predict future observations, there is a risk of overfitting the model.**

```{r}
# Create a random intercept and random slope model with the interaction:
BPRS_ref2 <- lmer(bprs ~ week + treatment + (week | subject) + (week * treatment), data = BPRSL, REML = FALSE)
summary(BPRS_ref2)
```

**Table 6.** Summary of *a random intercept and random slope model with the interaction of week and treatment*

```{r}
# Perform an ANOVA test on the two models:
anova(BPRS_ref2, BPRS_ref1)
```

**Table 7.** ANOVA test reveals that the interaction of week and treatment does not improve model fit in this BPRS data.

```{r, figures-side, fig.show="hold", out.width="50%"}
par(mar = c(4, 4, .1, .1))

# Draw the plot of BPRSL with the observed Weight values:
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(linetype = treatment, color = treatment)) +
  scale_x_continuous(name = "Weeks", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "BPRS", breaks = seq(0, 100, 20)) +
  theme(legend.position = "top")

# Create a vector of the fitted values (ref1):
Fitted <- fitted(BPRS_ref1)

# Create a new column fitted to RATSL:
BPRSL <- mutate(BPRSL, fitted = Fitted)

# Draw the plot of BPRSL with the Fitted values of weight:
ggplot(BPRSL, aes(x = week, y = fitted, group = subject)) +
  geom_line(aes(linetype = treatment, color = treatment)) +
  scale_x_continuous(name = "Weeks", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "BPRS", breaks = seq(0, 100, 20)) +
  theme(legend.position = "top")
```

**Figure 6.** The observed BPRS values on the left side, and the predicted BPRS values on the right hand side. In this figure I used *the random intercept and random slope model* for prediction since according to my previous interpretations that one works best in BPRS data.
