# WEEK 5: Dimensionality reduction techniques

In this exercise I was using data sets of human development and gender inequality. See the details from [this link](http://hdr.undp.org/en/content/human-development-index-hdi).

```{r, message=FALSE, warning=FALSE}
# Set working directory:
setwd("C:/Users/mikniemi/OneDrive - University of Helsinki/OpenDataScience/IODS/IODS-project")

# Access the necessary tool packages:
library(dplyr)
library(corrplot)
library(GGally)

# Load and explore the exercise data:
human <- read.table("Data/human.txt")
summary(human)
ggpairs(human)
```

There are in total 8 variables explaining human development: 1) **Percentage of females having secondary education**, 2) **Percentage of females being employed**, 3) **Expected years of schooling**, 4) **Expected life**, 5) **The gross national income (GNI) Index**, 6) **Maternal mortality**, 7) **Adolescent birth**, and 8) **Female representation in parliament**. The figure above illustrates correlations between each pair of variables.

```{r}
# Compute the correlation matrix and visualize it with corrplot:
cor(human) %>% corrplot(type = "upper", tl.cex=0.6)
```

This figure presents well, that variables 'Labour.Female' and 'Woman_rep.Parliament' have low correlations with all other variables, but otherwise the variables are correlating with each other quite much.



