# WEEK 4: Clustering and classification

In this exercise I used a data of **Housing Values in Suburbs of Boston** that was freely available in the MASS package. The data contains 506 rows (=different suburbs) and 14 columns (=variables). The variables try to explain different criteria of live quality in Boston suburbs. See the details [here](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html). 

```{r, message=FALSE, warning=FALSE}
# Access the necessary tool packages:
library(MASS)
library(corrplot)
library(dplyr)

# Load the Boston data:
data("Boston")

# Explore the data:
str(Boston)
summary(Boston)

# Calculate the correlation matrix and round it:
cor_matrix <- cor(Boston) %>% round(2)
cor_matrix
```

See the summary and the correlation matrix of the data above. For example we can notice that e.g. crime rate varies very much among the suburbs: the highest crime rate is even 89 %. I would also like to highlight large variations at air quality (nox = nitrogen oxides concentration), percent of lower status of the population (lstat), and proportion of blacks by town. Charles River (chas) is the only dummy variable there (= 1 if tract bounds river; 0 otherwise).

```{r}
# Visualize the correlation matrix:
corrplot(cor_matrix, method="circle", type="upper", cl.pos="b", tl.pos="d", tl.cex=0.6)
```

Here is the most beautiful illustration of correlation matrix that I have ever seen! You can copy the code above and use it, please! I don't believe the figure needs any explanation, since it's so good... :-)

### Standardizing the data set





